DESING AND SIMULATION SOFTWARE FOR RHODOTRON TYPE ELECTRON ACCELERATORS

### 2023-08-25 subsections/theory/rhodotron/rhodotron.tex
25/08/2023 update

@@

As particles gain energy in the accelerator, they approach relativistic speeds, where relativistic effects become significant. 
Special relativity governs the increase in mass and energy of the particles as they approach the speed of light, providing insights into the behavior of matter at high energies.
% TODO : ADD CIRCULAR AND LINEAR ACCELERATOR PICTURES HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{../../../figures/kahvelab_linac.png}
    \caption{A linear proton accelerator in KAHVELab \cite{kahvelab_linac}}
\end{figure}
\subsection{Key Concepts}

\subsubsection{Resonance Frequency}
Resonance frequency is the frequency in which the electromagnetic fields form standing waves in a cavity.
It is determined by the physical dimensions and the speed of light in the cavity's medium. 
In a rectangular cavity for example, the resonance frequency can be calculated from \fromeq{f_r_rec}.
\begin{equation}
    f_{klm} = \frac{c}{2\pi \sqrt{\mu_r \epsilon_r}} \sqrt{(\frac{k}{w})^2 + (\frac{l}{u})^2 + (\frac{m}{v})^2}
    \label{eq:f_r_rec}
\end{equation}
Where $w$, $u$, $v$ are the dimensions of the cavity, $\mu_r$ and $\epsilon_r$ are relative permability and permittivity of the cavity respectably.

@@

\subsubsection{Shunt Impedance}
The shunt impedance of an RF accelerator is a measure of the efficiency 
at which the accelerator can transform the supplied RF power into acceleration.

In addition to accelerating the particles, RF cavities are often designed to provide focusing forces. 
By carefully shaping the cavity and adjusting the electromagnetic fields, the particles can experience focusing effects as they pass through the cavity. 
This helps to maintain a tight and controlled beam. To ensure efficient acceleration, it is essential to maintain phase stability. 
This means that the particles should experience the strongest electric fields at the correct time during their passage through the cavity. 
Precise timing and synchronization of the RF power source with the particle beam are crucial to achieve phase stability and maximize energy transfer.
% TODO : ADD RF CAVITY PICTURE HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.75\textwidth]{../../../figures/pill_box.jpeg}
    \caption{An RF cavity used in KAHVELab}
\end{figure}

According to the Lorentz Force Law ( $\textit{Section \ref{sec:lorentz-force}}$ ), when a charged particle moves through a magnetic field, it experiences a force perpendicular to both its velocity 
vector and the magnetic field direction. This force causes the particle's trajectory to curve, resulting in a bending effect.

@@

Coaxial design of the cavity concentrates the electric field, while the magnetic field diminishes in the middle of the cylinders. 
Therefore the electrons are injected and accelerated in the plane of zero magnetic field where the electric field is strongest.
% TODO : ADD SUPERFISH AND CST SIMULATIONS OF A SIMPLE RHODOTRON CAVITY MAYBE?????

With the expectation that the electrons will accelerate to speeds $>=0.9 c$ after the first pass, a rhodotron cavity is designed so that the length of the path between successive passes is an integer multiple of $\lambda$, wavelength of the RF field ($l=p\lambda \label{eq:lpl}$).
This constraint helps with  phase stability and synchronization of the beam.

In the table below, optimized characteristics of a rhodotron cavity can be observed. 
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{../../../figures/pottier_table1.png}
    \caption{Optimized characteristics of a rhodotron cavity \cite{rhodo_pottier}}
    \label{fig:pottier_table1}
\end{figure}
Here, $p$ is the integer multiplier in the equation ($l=p\lambda$) mentioned above, $R_1$ is the radius of the inner cylinder, $R_2$ is the radius of the outer cylinder, 
$Z_{se}$ is effective shunt empedance, $Z_{sp}$ is practical shunt empedance which was taken to be $0.85 Z_{se}$.

Energy gain after each pass can be calculated by the following formula \cite{rhodo_pottier}
\begin{equation}
    \label{eq:W_gain_each_pass_pottier}
    W_1 = Z_{sp}^{1/2} P^{1/2} \cos \phi  
\end{equation}
where $P$ is the dissipated power, $\phi$ is the phase lag, typically taken as $15^\circ$ \cite{rhodo_pottier}.

Considering $Z_{sp} \propto \lambda^{1/2}$, \,\,\, $W_1 \propto \lambda^{1/4}$, \,\,\,   $V \propto \lambda^3$, where $V$ is the volume of the cavity, implementing the $p=1$ design in \fromfig{pottier_table1} is much more space efficient.
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{../../../figures/pottier_table2.png}
    \caption{Energy of a synchronous electron after each pass for both $p=1$ and $p=2$ \cite{rhodo_pottier}}
    \label{fig:pottier_table2}
\end{figure}
Total energy gain after $n$ passes, $W$, can then be found by \fromeq{W_gain_each_pass_pottier}, taking $\phi = 15^\circ$, $p=1$, i.e $R_2 = 0.27 \lambda$.
\begin{equation}
    \label{eq:W_total_gain_pottier}
    W \approx 2.14 \lambda^{1/4} P^{1/2} n \quad keV
\end{equation}

>>>>

\newcommand{\fromsec}[1]{\textit{section \ref{sec:#1}}}

@@

As particles gain energy in the accelerator, they approach relativistic speeds, where relativistic effects become significant. 
Special relativity governs the increase in mass and energy of the particles as they approach the speed of light, providing insights into the behavior of matter at high energies.
% TODO : ADD CIRCULAR AND LINEAR ACCELERATOR PICTURES HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{../../../figures/kahvelab_linac.png}
    \caption{A linear proton accelerator in KAHVELab \cite{kahvelab_linac}}
\end{figure}

In addition to accelerating the particles, RF cavities are often designed to provide focusing forces. 
By carefully shaping the cavity and adjusting the electromagnetic fields, the particles can experience focusing effects as they pass through the cavity. 
This helps to maintain a tight and controlled beam. To ensure efficient acceleration, it is essential to maintain phase stability. 
This means that the particles should experience the strongest electric fields at the correct time during their passage through the cavity. 
Precise timing and synchronization of the RF power source with the particle beam are crucial to achieve phase stability and maximize energy transfer.
\begin{figure}[H]
    \centering
    \includegraphics[width=.75\textwidth]{../../../figures/pill_box.jpeg}
    \caption{An RF cavity used in KAHVELab}
\end{figure}

According to the Lorentz Force Law (\fromsec{lorentz-force}), when a charged particle moves through a magnetic field, it experiences a force perpendicular to both its velocity 
vector and the magnetic field direction. This force causes the particle's trajectory to curve, resulting in a bending effect.

\subsection{Key Concepts}

\subsubsection{Resonance Frequency}
Resonance frequency is the frequency in which the electromagnetic fields form standing waves in a cavity.
It is determined by the physical dimensions and the speed of light in the cavity's medium. 
In a rectangular cavity for example, the resonance frequency can be calculated from \fromeq{f_r_rec}.
\begin{equation}
    f_{klm} = \frac{c}{2\pi \sqrt{\mu_r \epsilon_r}} \sqrt{(\frac{k}{w})^2 + (\frac{l}{u})^2 + (\frac{m}{v})^2}
    \label{eq:f_r_rec}
\end{equation}
where $w$, $u$, $v$ are the dimensions of the cavity, $\mu_r$ and $\epsilon_r$ are relative permability and permittivity of the cavity respectably.

@@

\subsubsection{Shunt Impedance}
The shunt impedance of an RF accelerator is a measure of the efficiency 
at which the accelerator can transform the supplied RF power into acceleration.
It is defined as
\begin{equation} \label{eq:shunt}
    Z_s = \frac{V_{acc}^2}{P_{diss}}
\end{equation}
where $V_{acc}$ is the accelerating potential in which the particle is subjected to, 
$P_{diss}$ is the power dissipated on the cavity walls. 
An example \textit{shunt impedance} calculation can be found in \fromsec{cavity_of_a_rhodotron}.

@@

Coaxial design of the cavity concentrates the electric field, while the magnetic field diminishes in the middle of the cylinders. 
Therefore the electrons are injected and accelerated in the plane of zero magnetic field where the electric field is strongest. (\fromfig{rhodo_cavity_field_dist})

\begin{figure}[H]
    \captionsetup[subfigure]{justification=centering}
    \captionsetup{justification=centering}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{../../../figures/cst/cavity_E_field_dist.png}
        \caption*{$\vecthreeBF{E}$ field.}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{../../../figures/cst/cavity_B_field_dist.png}
        \caption*{$\vecthreeBF{B}$ field.}
    \end{subfigure}
    \caption{$\vecthreeBF{E}$ and $\vecthreeBF{B}$ eigenmode field distributions inside a coaxial cavity.}
    \label{fig:rhodo_cavity_field_dist}
\end{figure}
For a cavity defined by the volume between two coaxial cylinders of equal lengths ($h$) with radii of $R_1$ and $R_2$, where $R_1 < R_2$, located at the origin (\fromfig{int_curve}),
first eigenmode solution of the $E$ and $B$ fields are \cite{rhodo_pottier}
\begin{eqnarray}
    E &=& \frac{E_0}{r} \cos(\frac{\pi z}{h}) \sin(\omega t + \phi) \\
    B &=& \frac{B_0}{r} \sin(\frac{\pi z}{h}) \cos(\omega t + \phi)
\end{eqnarray}
where $\omega=2\pi f$, $f$ is the resonance frequency.

\begin{figure}[H]
    \centering
    \includegraphics[width=.6\textwidth]{../../../figures/illustrations/rhodo_integral_curve.pdf}
    \caption{Illustration of a simple coaxial cavity with the curve \textbf{C} in \fromeq{p_diss_C_curve}}
    \label{fig:int_curve}
\end{figure}
Because acceleration potential is located on the $z=0$ plane and $\vecthreeBF{E} \parallel \hat{r}$, $V_{acc}$ can be found by the following equation:
\begin{eqnarray} 
    V_{acc} &=& 2 \int_{R_1}^{R_2} |E|^2 dr \\
            &=& 2 E_0 \int_{R_1}^{R_2} \frac{dr}{r} \\
            &=& 2 E_0 \ln(\frac{R_2}{R_1}) \label{eq:v_acc}
\end{eqnarray}
Dissipated power $P_{diss}$, on the other hand, can be calculated as follows \cite{rf}:
\begin{eqnarray} 
    P_{diss} &=& \frac{1}{2} \int \int \rho_s | H_{||} |^2 dA \nonumber \\
             &=& \frac{\rho_s}{2\mu_0^2} \int_0^{2\pi} \int_C | B_{||} |^2 r ds d\theta \label{eq:p_diss_C_curve}
\end{eqnarray}
where $\rho_s$ is the areal skin resistivity ( $\rho_s \approx 2.51 \times 10^{-7} f^{1/2}$ for copper \cite{rhodo_pottier} ), $B=\mu H$, $\mu$ is permeability, $\mu_0$ is the vacuum permability. 
Integral curve \textbf{C} is defined as the circumference of cylindrically symmetrical cross sectional area of the cavity (\fromfig{int_curve}).
This curve can be seperated to its line components and total power dissipation of this curve, $P_C$ will be equal to sum of the power dissipated in these lines.
Since $\vecthreeBF{B}$ is always parallel to the surface we can use $\vecthreeBF{B}$ directly:
\begin{eqnarray}
    P_{diss} &=& \frac{\rho_s}{2\mu_0^2} \int_0^{2\pi} (
                   2\int_{0}^{\frac{h}{2}} |B|^2 r dz \Big|_{r=R_1}
                 + 2\int_{0}^{\frac{h}{2}} |B|^2 r dz \Big|_{r=R_2} 
                 + 2\int_{R_1}^{R_2} |B|^2 r dr \Big|_{z=\frac{h}{2}})d\theta \nonumber\\
             %&=& \frac{\rho_s B_0^2 \cos^2(\omega t + \phi)}{2\mu_0^2} \int_0^{2\pi} 
             &=& \frac{\rho_s}{2\mu_0^2} \int_0^{2\pi} (2I_A + 2I_B + 2I_C)d\theta 
                 = \frac{2\rho_s \pi}{\mu_0^2}(I_A + I_B + I_C)
\end{eqnarray}
\begin{eqnarray}
    I_A = \int_{0}^{\frac{h}{2}} |B|^2 r dz \Big|_{r=R_1} &=& B_0^2 \frac{h}{4R_1} \\
    I_B = \int_{0}^{\frac{h}{2}} |B|^2 r dz \Big|_{r=R_2} &=& B_0^2 \frac{h}{4R_2} \\
    I_C = \int_{R_1}^{R_2} |B|^2 r dr \Big|_{z=\frac{h}{2}} &=& B_0^2 \ln(\frac{R_2}{R_1})
\end{eqnarray}
inserting $H_0=B_0/\mu_0$, finally we have the dissipated power:
\begin{equation} \label{eq:p_diss}
    P_{diss} = \rho_s \pi H_o ( \frac{h}{2R_1} + \frac{h}{2R_2} + 2\ln(\frac{R_2}{R_1}) )
\end{equation}
Therefore, from \fromeq{shunt}, using \fromeqs{v_acc}{p_diss} also $E_0/H_0 = Z_0 \approx 120 \pi$:
\begin{eqnarray}
    Z_s &=& \frac{4E_0^2}{H_0^2 \pi \rho_s} \frac{ \ln^2(\frac{R_2}{R_1})}{(\frac{h}{2R_1} + \frac{h}{2R_2} + 2 \ln(\frac{R_2}{R_1}))} \\
        &=& \frac{8 \pi 60^2}{\rho_s} \frac{ \ln^2(\frac{R_2}{R_1})}{(\frac{h}{4}(\frac{1}{R_1} + \frac{1}{R_2}) + \ln(\frac{R_2}{R_1}))}
\end{eqnarray}
where, time dependencies have been removed from $\vecthreeBF{E}$ and $\vecthreeBF{B}$ 
and the maximum values for $V_{acc}$ and $P_{diss}$ have been used. 
However, particles do not interact with constant $\vecthreeBF{E}$ field during acceleration. 
Therefore, a more useful parameter called \textit{effective shunt impedance} can be defined:
\begin{equation}
    Z_{se} = Z_s T^2
\end{equation}
where $T$ is the \textit{transit time factor}, a correctional coefficent that contain the changing field effects during acceleration.
For a relativistic electron crossing the axis at time 0, $T$ can be found by \cite{rhodo_pottier}:
\begin{eqnarray}
    T &=& \frac{S_i(\frac{2 \pi R_2}{\lambda}) - S_i(\frac{2 \pi R_1}{\lambda})}{\ln(\frac{R_2}{R_1})} \\
    S_i(\theta) &=& \int_0^{\theta} \frac{\sin(x)}{x} dx
\end{eqnarray}
for a relativistic electron crossing the axis at $2\pi f t_0 = \phi$ on the other hand, $T$ needs to be multiplied by $\cos(\phi)$:
\begin{equation}
    T(\phi) = T \cos(\phi)
\end{equation} 
Putting all these calculations together, energy gain of a relativistic electron, passing the origin at $\phi / \omega$ as calculated by \textit{J. Pottier} is \cite{rhodo_pottier}
\begin{eqnarray}
    \Delta E &=& q V_{acc}^{ef} \\
    \Delta E &=& q Z_{se}^{1/2}P_{diss}^{1/2} \cos(\phi)
\end{eqnarray}
where $ V_{acc}^{ef} = V_{acc} T(\phi)$ is the effective accelerating potential.
If $\Delta E$ is taken in \textit{MeV}, $Z_{se}$ in $M\Omega$ and $P_{diss}$ in \textit{MW}, this equality becomes
\begin{equation} \label{eq:W_gain_each_pass_pottier}
    \Delta E = Z_{se}^{1/2}P_{diss}^{1/2} \cos(\phi) \quad MeV
\end{equation}
With the expectation that the electrons will accelerate to speeds $\approx c$ after the first pass, a rhodotron cavity is designed so that the length of the path between successive passes is an integer multiple of $\lambda$, wavelength of the RF field ($l=p\lambda \label{eq:lpl}$).
This constraint helps with  phase stability and synchronization of the beam.

In the table below, optimized characteristics of a rhodotron cavity can be observed. 
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{../../../figures/pottier_table1.png}
    \caption{Optimized characteristics of a rhodotron cavity \cite{rhodo_pottier}}
    \label{fig:pottier_table1}
\end{figure}
Here, $p$ is the integer multiplier in the equation ($l=p\lambda$) mentioned above, $R_1$ is the radius of the inner cylinder, $R_2$ is the radius of the outer cylinder, 
$Z_{se}$ is effective shunt empedance, $Z_{sp}$ is practical shunt empedance which was taken to be $0.85 Z_{se}$. Typically, phase lag $\phi$ is taken as $15^\circ$ \cite{rhodo_pottier}.

Considering $Z_{sp} \propto \lambda^{1/2}$, \,\,\, $\Delta E \propto \lambda^{1/4}$, \,\,\,   $V \propto \lambda^3$, where $V$ is the volume of the cavity, implementing the $p=1$ design in \fromfig{pottier_table1} is much more space efficient.
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{../../../figures/pottier_table2.png}
    \caption{Energy of a synchronous electron after each pass for both $p=1$ and $p=2$ \cite{rhodo_pottier}}
    \label{fig:pottier_table2}
\end{figure}
Total energy gain after $n$ passes, $\Delta E_n$, can then be found by \fromeq{W_gain_each_pass_pottier}, taking $\phi = 15^\circ$, $p=1$, i.e $R_2 = 0.27 \lambda$, $P$ in \textit{W}, $\lambda$ in \textit{m}.
\begin{equation}
    \label{eq:W_total_gain_pottier}
    \Delta E_n \approx 2.14 \lambda^{1/4} P^{1/2} n \quad keV
\end{equation}

### 2023-08-25 subsections/tools/available/available.tex
25/08/2023 update

\subsection{CST Studio Suite}
CST Studio Suite is a powerful software package developed by \textit{Simulia} for electromagnetic simulation and analysis. It's widely used in various industries, including electronics, telecommunications, automotive, aerospace, and more, to design and optimize electromagnetic devices and systems. 
The software provides a comprehensive set of tools for simulating and analyzing the behavior of electromagnetic fields and their interactions with different materials, structures and charged particles.

\subsection{ROOT}
ROOT is a widely used open-source software framework developed by CERN (European Organization for Nuclear Research) for data analysis,
visualization, and storage in the field of high-energy physics, particularly in experiments conducted at particle accelerators 
like the Large Hadron Collider (LHC) \cite{root}. It has become an essential tool in particle physics research, 
and it's used by physicists around the world to analyze and interpret data.

>>>>

\subsection{CST Studio Suite}
CST Studio Suite is a powerful software package developed by \textit{Simulia} for electromagnetic simulation and analysis. It is widely used in various industries, including electronics, telecommunications, automotive, aerospace, and more, to design and optimize electromagnetic devices and systems. 
The software provides a comprehensive set of tools for simulating and analyzing the behavior of electromagnetic fields and their interactions with different materials, structures and charged particles.

\subsection{ROOT}
ROOT is a widely used open-source software framework developed by CERN (European Organization for Nuclear Research) for data analysis,
visualization, and storage in the field of high-energy physics, particularly in experiments conducted at particle accelerators 
like the Large Hadron Collider (LHC) \cite{root}. It has become an essential tool in particle physics research, 
and it is used by physicists around the world to analyze and interpret data.
